---
title: "Getting help with R"
output:
        html_document:
            includes:
                in_header: header.html
            theme: paper
            toc: true
            toc_depth: 4
            toc_float: true
            number_sections: false
            fig_caption: false
            markdown: GFM
            code_download: false
---

<style type="text/css">

body, td {
   font-size: 16px;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 12px
}

</style>

```{r, include = FALSE}
source("../bin/set_values.R")
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(lang = c("r", "markdown", "bash"), position = c("top", "right"))
```

```{r, 'chunk_options', include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")
```

# Objectives

- Learn where to get help.
- Learn how to ask for help with R.

# Finding help

Part of trying new thinghs is occasionally getting stuck and needing help. 
Fortunately, you're not alone, and if you can't find a previous example of how
to make something work, you can ask for help from the community of R folks. Here
are some tips to make this process as helpful and efficient as possible.

- R's own [suggestions in getting help](https://www.r-project.org/help.html){target="_blank"} are excellent.
- Google is your friend and AI can also be your friend. But on occasion, they 
will cheerfully suggest a *completely wrong answer*, so **always verify any 
proposed solution actually works for your problem**. 

There are a few websites that are useful for getting help with R.

| Website | Purpose |
| :--- | :--- |
| [R Documentation](https://www.rdocumentation.org/){target="_blank"} | Searchable documentation for R packages on CRAN and Bioconductor |
| [Bioconductor Support](https://support.bioconductor.org/){target="_blank"} | Support forum for Bioconductor packages |
| [stackoverflow](https://stackoverflow.com/questions/tagged/r){target="_blank"} | General programming questions |
| [Biostars](https://www.biostars.org/){target="_blank"} | Bioinformatics specific questions |
| [R for data science](https://r4ds.hadley.nz/){target="_blank"} | An extremely readable website book that goes soup to nuts. |
| [RWeekly](https://rweekly.org/){target="_blank"} | Short kind of random tips/tidbits updated weekly |
| [Posit Community Forum](https://forum.posit.co){target="_blank"} | Questions asked and answered |
| [Posit: Cheatsheets](https://posit.co/resources/cheatsheets/){target="_blank"} | Super handy since no one has all the packages memorized |
| [Posit: Getting help with R](https://support.posit.co/hc/en-us/articles/200552336-Getting-Help-with-R){target="_blank"} | Curated links by the posit team |

# How to ask for help

If you can't find the right solution already posted, you you can post your new question to a forum.
R's [Posting Guide](https://www.r-project.org/posting-guide.html){target="_blank} examples of how to ask for help with R. 
It's a good read - but here's the condensed version:

1. **Re-check if a question has already been asked**. &#x1F60E; 
2. **Clearly describe your problem in words**.  
  State what you're trying to do, what you tried, and the result, including the full text of any error message.
3. **Try to include a "minimally reproducible example"**.  
  This means providing the smallest block of data and code that anyone can run to 
  reproduce the error and (we hope) diagnose the issue. Focusing the question on 
  the suspect code with minimal data helps the respondents dive right in. (Often '
  times, when you're trying to craft the minimal example, you learn something new 
  and sometimes even resolve the original problem.)

  ```{verbatim, minimal_example, eval=FALSE}
  I have this data frame:
  
  my_data = read_csv("start, end, title
  8:00, 4:00, Using R/R Studio
  1:00, 4:00, Nextflow: Train the Trainers
  8:00, 4:00, Lean Management for Shared Resources
  ")
  
  But I want this data frame:
  title                                workshop time  
  Using R/R Studio                     start    08:00 
  Using R/R Studio                     end      04:00 
  Nextflow: Train the Trainers         start    01:00 
  Nextflow: Train the Trainers         end      04:00 
  Lean Management for Shared Resources start    08:00 
  Lean Management for Shared Resources end      04:00
  
  ```
  Note that sharing **minimal example data** often means you need to trim the data down to a small subset
   
   - Shown in the block above, `read_csv` allows you to "handcraft" data as a comma separated list of fields, one row per field.
   - You can use one of R's built in example datasets like `iris` or `mtcars`. You can see the full list of built in datasets with `data()`.
   - You can also use `format_csv` and `head` to prune a larger dataset to a few rows, e.g. `cat(format_csv(head(mtcars)))` to then 
     copy into your post.

4. **Share your session info**.  
  Your problem and the solution often depends on the specific package versions you are using, so you should share this in your post.
  Sharing help others reproduce the issue and find a solution quicker. There are a couple ways to get the session information. The older/more common way is the `sessionInfo` command which will work in any R environment:

```{r eval = exists("on_aws") && on_aws, echo=FALSE}
# To avoid confusion, we will only generate/refresh these files on AWS instance
# The on_aws var is initialized in bin/set_values.R

sink('sessionInfo.txt')
sessionInfo()
sink()

sink('session_info.txt')
sessioninfo::session_info()
sink()
```


```{r sessionInfo, eval=FALSE}
################################################################################
# Print out details about this R install, session, and loaded libraries
sessionInfo()
```

```{verbatim sessionInfo_txt, file='sessionInfo.txt', eval=FALSE}
```

The new, slightly more elegant is `sessioninfo::session_info()`. (Note,
the `sessioninfo` package must be installed. We did that for you.)


```{r session_info, eval=FALSE}
################################################################################
# A more concise, pleasant version of sessionInfo()
# install.packages('session_info')
sessioninfo::session_info()
```

```{verbatim session_info_txt, file='session_info.txt', eval=FALSE}
```

<br/>

"All of us, at some time or other, need help. Whether we're giving or receiving help, each one of us has something valuable to bring to this world."  
— Fred Rogers

<br/>
<hr/>
| [Previous lesson](r-04-bioconductor.html) | [Top of this lesson](#top) | [Next lesson](r-03-exploration-ggplot.html) |
| :--- | :----: | ---: |