---
title: "Wrapping up"
output:
        html_document:
            includes:
                in_header: header.html
            theme: paper
            toc: true
            toc_depth: 4
            toc_float: true
            number_sections: false
            fig_caption: true
            markdown: GFM
            code_download: false
---

<style type="text/css">

body, td {
   font-size: 18px;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 12px
}

</style>

```{r, include = FALSE}
source("../bin/set_values.R")
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(lang = c("r", "markdown", "bash"), position = c("top", "right"))
```

We hope you now have more familiarity with key concepts, tools, and techniques
that will enable more efficient, reproducible, and impactful data research and 
science.


## Session Info

<details>
```{r session_info, eval=FALSE}
################################################################################
# Session info (for reproducibility / future reference)
session_info::session_info()
```

```{verbatim session_info_txt, file='session_info.txt', eval=FALSE}
```
</details>


---

## Housekeeping

- This workshop website (aka "show notes") will be available.

---

### Can I continue to use the RStudio environment we used in the workshop?
- The [PositCloud](https://posit.cloud/){target="_blank"} environment used in this workshop does not expire. 
- You have the option of downloading the workshop files to your own computer
  ```
  mkdir workshop-intro-r-rstudio
  scp -r YOUR_USERNAME@`r workshop_vars$ssh_download_dns`:"IRR*" workshop-intro-r-rstudio
  ```
### Where else could I use RStudio?
  - See [Advanced setup instructions](workshop_setup/setup_instructions_advanced.html){target="_blank"}
    for details on how to install RStudio and required packages on your own computer.

---

### Where can I learn more about RStudio packages and functions?
- A nifty collection of [cheatsheets for R](https://www.rstudio.com/resources/cheatsheets/){target="_blank"} and links to download some our favorites:
  - [Base R](http://github.com/rstudio/cheatsheets/blob/main/base-r.pdf){target="_blank"}   
  - [Data transformation with dplyr](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf){target="_blank"}
  - [Data visualization with ggplot2](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf){target="_blank"}

---

### Can you recommend other relevant workshops or tutorials?
- The [University of Michigan Bioinformatics Core](https://michmed.org/GqGzZ){target="_blank"} regularly hosts workshops.
- Intro lessons and workshops in Bash / Git / R / Python : 
  - [Software Carpentry](https://software-carpentry.org/lessons/){target="_blank"}
- Some great "practice" datasets 
  - [tidytuesday](https://github.com/rfordatascience/tidytuesday){target="_blank"} : a site for sharing sample datasets
  - [Kaggle](https://www.kaggle.com/datasets){target="_blank"} : a site for learning AI/ML, but you can jump right to the sample datasets &#128521;

## Thank you

Thank you for participating in our workshop. We welcome your questions and
feedback now and in the future.

The Workshop Team

[bioinformatics-workshops@umich.edu](mailto:bioinformatics-workshops@umich.edu) <br/>
[UM BRCF Bioinformatics Core](https://medresearch.umich.edu/office-research/about-office-research/biomedical-research-core-facilities/bioinformatics-core){target="_blank"}

<hr/>